Boosted Regression Trees for Data-Driven Auto-Tuning

### 2011-12-07 paper/draft1.tex
m

\title{Boosted Regression Trees for Model-Free Auto-Tuning}

@@

\begin{abstract}
Autotuning is a widely used and effective technique for optimizing a
parametrized GPU code template for a particular type of computation on
particular hardware.
Its drawback is that thorough or exhaustive auto-tuning can be very slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations with a wide variety of memory
access patterns and computational bottlenecks.
How can we draw on data from previous auto-tuning of related problems on related hardware to make a
good decision on a novel problem?
This paper presents a machine learning approach to auto-tuning, in
which features of
(a) the current hardware platform,
(b) the kernel configuration
and (c) the problem instance are used to fit a regression model (boosted regression trees) that predicts
how much faster an implementation will be than a reference baseline.
Combinatorial optimization strategies that would normally be applied for auto-tuning can be applied to the regression model to find an implementation much more quickly.
We validate our approach using the filterbank correlation kernel described in [anon].
We find that 0.05 seconds of hill climbing on the regression model can achieve
80\% of the improvement brought by 120 seconds of real auto-tuning.
Our approach is not specific to filterbank correlation, or even to GPU auto-tuning: the approach of using a non-linear regression model on top of simple features applies to a variety of problem types, kernel types, and platforms.
A detailed platform model can provide useful features for
cross-platform auto-tuning, but such a model is not necessary for same-platform autotuning.
\end{abstract}

@@

\subsection{Problem Space: Filter-bank Correlation}

Our experiments on automatic auto-tuning will be focused on filterbank
correlation.  Mathematically, we define filterbank correlation in terms of an
image $x$ and a filterbank $f$.
The image $x$ has $R$ rows, $C$ columns, and $D$ channels (e.g. color
channels) that we call its {\em depth}. We index $x$ like $\mathbf{x}[i,j,d]$
where $1 \leq i \leq R$ and $1 \leq j \leq C$.
The filterbank $f$ has $K$ square filters of side-length $W$, each one of
which has $D$ channels.
The result of filterbank correlation of $x$ with $f$ is an image-like array
$z$ with $R-W+1$ rows, $C-W+1$ columns, and depth $K$, whose elements are
defined according to Equation~\ref{eq:z}:

\begin{equation}
    \mathbf{z}[i,j,k] = \sum_{u=1}^{W} \sum_{v=1}^{W} \sum_{d=1}^D
        \mathbf{x}[i+u, j+v, d]~ \mathbf{f}[k, u, v, d].
        \label{eq:z}
\end{equation}

Filterbank correlation is used extensively in image and video processing,
where it is often the computational bottleneck.
GPUs are well-suited to computing filterbank correlations, although the
optimal strategy for doing so often depends on the shape of the problem, i.e.
the choices of $R$, $C$, $D$, $F$, and $W$.

We randomly sampled problem specifications from the following space:
\begin{figure}
\begin{center}
\begin{minipage}{.4\textwidth}
\begin{verbatim}
R : one_of(256, 512, 1024, 2048, 4096)
D : one_of(1, 4, 8, 16, 32, 64, 128, 256)
F : one_of(1, 4, 8, 16, 32, 64, 128, 256)
W : one_of(3, 5, 7, 9, 11)
\end{verbatim}
\end{minipage}
\end{center}
\caption{The full configuration space for filterbank correlation included a
total of 1600 elements, but we restricted our experiments to problems that
represented between 1 and 50 gigaflops. Our problem space included 602
configurations. In a full application in which all image and filter sizes are
valid, the number of configurations would be effectively infinite.
XXX: define R, D, F, W here
}
\label{fig:probspace}
\end{figure}
% Problems were considered invalid if the filters were larger than the images.
% but in the new setting with iheight >= 256 this never happened.
Problems were tested if they represented between 1 and 50 gigaflops.  Problems
that were smaller do not fully utilize the hardware, and are handled equally
well by many kernel settings, including the reference (XXX cite GCG).
Problems that were larger take so long to evaluate that there is negligible
inefficiency in implementing them via multiple calls with smaller images and fewer filters.
The configuration space described in Figure~\ref{fig:probspace} included 602
configurations with between 1 and 50 gigaflops.

@@

\begin{figure}
\begin{center}
\begin{minipage}{.43\textwidth}
\begin{verbatim}
block_w   : one_of(4, 8, 16, 32, 64, 128)
block_h   : one_of(4, 8, 16, 32, 64, 128)
n_filter_r: one_of(1, 2)
n_output4s: one_of("all", 1, 2)
spill     : one_of(False, True)
imul_fast : one_of(False, True)
pad_shared: one_of(False, True)
use_tex1d : one_of(False, True)
maxrreg   : one_of(8, 16, 20, 24, 28, 32, inf)
fast_math : one_of(False, True)
\end{verbatim}
\end{minipage}
\end{center}
\caption{The kernel used to perform
filterbank correlation was parametrized by 10 parameters, some of which
were binary and others of which were integer-valued. The full configuration
space included 12000 kernels.
XXX: how to make sense of these paramters without code listing?
}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[scale=.42]{speedup_295.pdf}
\includegraphics[scale=.42]{speedup_480.pdf}
\caption{Speedup of various auto-tuning strategies over a good reference
implementation.
Each strategy was evaluated the same set of 83 problem configurations, which
was disjoint from the problem configurations used to build the model for
model-based hill-climbing.
The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
of each strategy. The grid and hill-climbing approaches
tested 73 and 75 configurations respectively, and took 45?? XXX
seconds on average.
The model-based approach tested 0 configuration evaluations, and took 0.05
seconds on average, even with a naive Python implementation.
}
\label{fig:speedup_295}
\end{figure*}

@@

>>>>

\title{Boosted Regression Trees for Data-Driven Auto-Tuning}

@@

\begin{abstract}
Auto-tuning is a widely used and effective technique for optimizing a
parametrized GPU code template for a particular computation on particular
hardware.  Its drawback is that thorough or exhaustive auto-tuning requires
compiling many kernels and calling each one many times; this process is slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations with a wide variety of memory access
patterns and computational bottlenecks.  How can we draw on data from previous
empirical auto-tuning of related problems on related hardware to make a just-in-time
implementation decision for a novel problem?  This paper presents a machine
learning approach to auto-tuning, in which features of (a) the current
hardware platform, (b) the kernel configuration and (c) the problem instance
are passed to a regression model (boosted regression trees) which predicts how
much faster this kernel will be than a reference baseline.  Combinatorial
optimization strategies that would normally implement auto-tuning by
evaluating kernel configurations on the real hardware are orders of magnitude
faster when evaluating the regression model instead.  We validate our approach
using the filterbank correlation kernel described in [XXX], where we find
that 0.1 seconds (XXX) of hill climbing on the regression model can achieve an
average of 95\% of the speed brought by 120 seconds of true auto-tuning XXX.  Our
approach is not specific to filterbank correlation, or even to GPU
auto-tuning: the approach of using a non-linear regression model on top of
simple features applies to a variety of problem types, kernel types, and
platforms.
\end{abstract}

@@

\section{Filter-bank Correlation}

Following [XXX], the experiments in this paper will focus on filterbank
correlation.  Mathematically, we define filterbank correlation in terms of an
image $x$ and a filterbank $f$.
The image $x$ has $R$ rows, $C$ columns, and $D$ channels (e.g. color
channels) that we call its {\em depth}. We index $x$ like $\mathbf{x}[i,j,d]$
where $0 \leq i < R$, $0 \leq j < C$, and $0 \leq d < D$.
The filterbank $f$ has $F$ filters that are like little images: each has a
height $H$, a width $W$, and $D$ channels.
We will restrict ourselves to what are called {\em valid} correlations, in
which the image is larger in both rows and columns than the filters.
The result of filterbank correlation of $x$ with $f$ is an image-like array
$z$ with $R-H+1$ rows, $C-W+1$ columns, and depth $F$, whose elements are
defined according to Equation~\ref{eq:z}:

\begin{equation}
    \mathbf{z}[r,c,k] = \sum_{w=0}^{W-1} \sum_{h=0}^{H-1} \sum_{d=0}^{D-1}
        \mathbf{x}[r+h, c+h, d]~ \mathbf{f}[k, h, w, d].
        \label{eq:z}
\end{equation}

Filterbank correlation is used extensively in image and video processing,
where it is often the computational bottleneck.
GPUs are well-suited to computing filterbank correlations, although the
optimal strategy for doing so often depends on the shape of the problem, i.e.
the choices of $R$, $C$, $D$, $F$, and $W$.
The computation of filterbank correlation on GPUs was recently treated by
XXX, which serves as the basis for the present work.

\section{Problem Configuration}

We randomly sampled problem specifications from the following space:
\begin{align*}
R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
H = W & \in \{ 3, 5, 7, 9, 11 \} \\
D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
\end{align*}
The problem configuration space used in our experiments. The product space
includes 1600 elements, but we restricted our experiments to problems that
represented between 1 and 50 gigaflops. Our problem space included 602
configurations. In a full application in which all image and filter sizes are
valid, the number of configurations would be effectively infinite.

% Problems were considered invalid if the filters were larger than the images.
% but in the new setting with iheight >= 256 this never happened.
Problems were tested if they represented between 1 and 50 gigaflops.  Problems
that were smaller do not fully utilize the hardware, and are handled equally
well by many kernel settings, including the reference (XXX cite GCG).
Problems that were larger take so long to evaluate that there is negligible
inefficiency in implementing them via multiple calls with smaller images and fewer filters.
The configuration space described in Figure~\ref{fig:probspace} included 602
configurations with between 1 and 50 gigaflops.

@@

\begin{align*}
\mathrm{block\_w}    & \in (4, 8, 16, 32, 64, 128) \\
\mathrm{block\_h}    & \in (4, 8, 16, 32, 64, 128) \\
\mathrm{n\_filter\_r} & \in (1, 2) \\
\mathrm{n\_output4s} & \in (\mathrm{all}, 1, 2) \\
\mathrm{spill}      & \in (False, True) \\
\mathrm{imul\_fast}  & \in (False, True) \\
\mathrm{pad\_shared} & \in (False, True) \\
\mathrm{use\_tex1d}  & \in (False, True) \\
\mathrm{maxrreg}    & \in (8, 16, 20, 24, 28, 32, \infty) \\
\mathrm{fast\_math}  & \in (False, True)
\end{align*}

The kernel used to perform
filterbank correlation was parametrized by 10 parameters, some of which
were binary and others of which were integer-valued. The full configuration
space included 12000 kernels.
XXX: how to make sense of these paramters without code listing?

@@

\begin{figure*}
\centering
\includegraphics[scale=.42]{speedup_295.pdf}
\includegraphics[scale=.42]{speedup_480.pdf}
\caption{Speedup of various auto-tuning strategies over a good reference
implementation.
Each strategy was evaluated the same set of 83 problem configurations, which
was disjoint from the problem configurations used to build the model for
model-based hill-climbing.
The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
of each strategy. The grid and hill-climbing approaches
tested 73 and 75 configurations respectively, and took 45?? XXX
seconds on average.
The model-based approach tested 0 configuration evaluations, and took 0.05
seconds on average, even with a naive Python implementation.
}
\label{fig:speedup}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[scale=.42]{fig_ntrain_295.pdf}
\includegraphics[scale=.42]{fig_ntrain_480.pdf}
\caption{The effect of the invalid configuration score ($a$) and training set
size on simulated auto-tuning.
All candidates timed during the grid and
hill-climbing search procedures were used as training points, so the training
set sizes ranged from an average of 1500 (10 problem configurations) to 30
thousand (200 problem configurations).
Training on 10 or 25 configurations was useful (higher mean than the reference),
but not as useful as training on 50 or more configurations.
The results on the GTX 295 suggest that
moderate values of $a$ between $0.25$ and $0.75$ might be
best, but $a$ had no significant effect on the GTX 480.
}
\label{fig:fig_ntrain}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[scale=.42]{fig_gflop_scatter_295.pdf}
\includegraphics[scale=.42]{fig_gflop_scatter_480.pdf}
\caption{Computation speed for novel problem configurations using predictive
vs. empirical code specialization.
The scatterplots are roughly symmetric about the axis of equality (main diagonal)
with occasional outliers, indicating that for most problem configurations the
predictive approach gives as good a solution as the empirical approach.
The predictive approach typically requires about 0.1 seconds to suggest an
implementation, whereas the empirical approach requires about 2 minutes.
}
\label{fig:fig_gflop_scatter}
\end{figure*}

@@

In a complete implementation of Eq.~\ref{eq:z}, there would be parameters
related to the physical layout (e.g. strides) of $x$, $f$ and $z$ arrays in
addition to the mathematical parameters of heights and widths and so on,
so the total number of filterbank correlation computations that a user might
be able to demand is astronomical.
A typical coping mechanism would be to cast an arbitrary problem configuration into a more
standardized form, such as by copying inputs into aligned memory buffers with
appropriate padding, and then choosing an appropriate blocking strategy for
the computation. At that point, auto-tuning efforts can be focussed on the
kernel for each blocked strategy. However, on GPU hardware, the cost of
aligning the inputs can be relatively large.  In future work we hope to apply
our model-based technique to auto-tune in full problem configuration space, so
that we can optimize as much of the computation as possible within the
predictive auto-tuning framework (XXX name).
